---
title: Predicting Eastern Indigo Snake Occurrences in Florida Using Construction Permitting
  Data
author: "Jonathan Bunch"
date: "8/6/2020"
output:
  pdf_document: default
  html_document: default
bibliography: bunch_indigo_bibliography.bib
nocite: '@*'
---

```{r setup, include=FALSE}
library(ggplot2)
library(readxl)
library(plyr)
library(stats)
library(gvlma)
```

# Introduction
The conservation of threatened species is an important, albeit daunting, endeavor.
Unfortunately, many ecologically important species are overlooked by the general 
public simply because they are not considered appealing.  Snake conservation is 
particularly difficult to promote, due to the common fears and misconceptions 
associated with them.  The Eastern Indigo snake (Drymarchon couperi) is a large, 
non-venomous snake native to the south-eastern United States.  This species has 
great value as a predator in the ecosystems it inhabits, feeding on many common 
pest animals.  This species also has the unique ability to predate venomous snake 
species, helping to control the population of species that are potentially dangerous 
to humans.  

The Eastern Indigo snake is currently listed as “threatened” by the US Fish and 
Wildlife Service.  The primary cause for declining populations is thought to be 
habitat destruction and fragmentation.  These snakes are known to co-occupy the 
borrows of gopher tortoises (Gopherus polyphemus), whose population has also sharply 
declined, adding further difficulties for the species.  While both of these species 
are legally protected, it is rumored that loopholes are commonly exploited by real-
estate developers to bypass mandatory reporting and damage mitigation processes.

The goal of this project was to explore the impact of new construction projects on
the observed occurrences of this species in Florida.  Observed occurrences (sightings) 
of this species are documented and verified by several scientific organizations and 
made available as an organized data set.  New construction data were collected from
the U.S. Census Building Permits Survey.


# Problem Statement
Habitat destruction and fragmentation is a leading cause of species decline, and
human habitation is a leading cause of habitat destruction and fragmentation.  
Historically, humans simply did not understand the impact that they were having on
their environment.  Over time we have become increasingly aware of the profound
impact our actions have on the natural world, and the importance of taking those
impacts into consideration.  The more clearly and specifically we can quantify our
most harmful practices, the better positioned we will be to change industry standards
and create effective conservation legislation.  This project will explore different
types of residential housing, and different measures of quantity of new housing,
to search for specific indicators of species decline.

# Methods
Two data sets were collected and prepared for this project: an observed occurrences
of rare species data set and a yearly permit survey data set.  Unnecessary and irrelevant
variables were removed and the data sets were conditionally restructured to match
the desired species, time frame, and location.  These decisions were based on several
considerations, including missing data and number of observations.  Once the basic
structure was established I began exploratory analysis.

```{r}
#### Occurrences Data Preliminary Preparation
# Import data and create subsets based on values of relevant variables.
occurrences <- read.delim("/Users/jonathanbunch/Documents/dsc520/final-project/data/drymarchon_occ/occurrence.txt")
occ_sub1 <- subset(occurrences, select = c(basisOfRecord, year, stateProvince, 
                                           specificEpithet, issue))
occ_sub2 <- subset(occ_sub1, stateProvince == "Florida" & specificEpithet == "couperi" 
                   & year >= 1970 & basisOfRecord == "HUMAN_OBSERVATION")
# Plot to see the distribution of sightings.
ggplot(data = occ_sub2, aes(x = year)) + geom_histogram(bins = 50) + 
  ggtitle("Occurrence Data Set Histogram") + xlab("Year") + ylab("Count")

#### Restructure the Occurrences Data to Represent Sightings per Year
# Create a new data table that counts the sightings per year.
occ_count <- as.data.frame(table(occ_sub2[, "year"]))
# This method converted the year to a factor. This will convert it back to a number.
occ_count$Var1 <- as.numeric(as.character(occ_count$Var1))
# The year 2015 had no occurrences, so it was not included in our count data frame.
# We need to add it to the data frame with a zero value for frequency.
occ_count <- rbind(occ_count, c(2015, 0))
# The order() function will move the new row to the correct chronological position.
occ_count <- occ_count[order(occ_count$Var1), ]
# Plot to make sure the "sightings per year" data frame matches the histogram.
ggplot(data = occ_count, aes(x = Var1, y = Freq)) + geom_col() + 
  ggtitle("Yearly Observed Occurrences") + xlab("Year") + ylab("Occurrences")

#### Permitting Data Preliminary Preparation
# Import data and create subsets based on values of relevant variables.
permits <- read_xlsx("/Users/jonathanbunch/Documents/dsc520/final-project/data/fl_permits.xlsx")
permits$year <- as.integer(permits$year)
permits_sub1 <- subset(permits, select = -c(`1_unit_buildings`, `2_unit_buildings`,
                                            `3_and_4_unit_buildings`, `5_plus_unit_buildings`,
                                            total_buildings))
permits_sub2 <- subset(permits_sub1, year >= 1970)
permits_sub2$year <- as.integer(permits_sub2$year)
# These final two variables were selected based on their correlation with the occurrences data.
permits_sub3 <- subset(permits_sub2, select = c(total_construction_valuation, `1_unit_valuation`))
```

Analysis began with trial and error experimentation through restructuring, visualization, 
and correlation testing.  These tests lead to further exclusions of variables that
did not appear to have statistical significance in this case.  

```{r}
cor(x = occ_count$Freq, y = permits_sub2)
```

Due to the low sample size of the occurrences data set, I chose to aggregate the 
yearly sighting count into five year blocks.  The two most promising variables from 
the permitting data set were aggregated into the same five year blocks and combined 
with the occurrences data to produce my final data set.

```{r}
#### Aggregating Yearly Values into Five Year Totals.
# This function returns the sum of each five row block for a given variable.
blocks_func <- function(x) {
  return(c(sum(x[1:5]), sum(x[6:10]), sum(x[11:15]), sum(x[16:20]), sum(x[21:25]), 
           sum(x[26:30]), sum(x[31:35]), sum(x[36:40]), sum(x[41:45]), sum(x[46:50])))
}
# Labels for each block of years.
block_labels <- c("1970-1974", "1975-1979", "1980-1984", "1985-1989", "1990-1994",
                 "1995-1999", "2000-2004", "2005-2009", "2010-2014", "2015-2019")
# New data frame with labels and sums of values for each five year period.
blocks_df <- data.frame(years = block_labels, occurrences = blocks_func(occ_count$Freq),
                        total_const_valuation = blocks_func(permits_sub3$total_construction_valuation),
                        one_unit_valuation = blocks_func(permits_sub3$`1_unit_valuation`))
# See what the five year block data frame looks like.
head(blocks_df)
# Correlation matrix.
cor(blocks_df[,-1])
```

Once the data were prepared, I performed simple and multiple regression to create
models of observed occurrences based on the two permitting variables.  This resulted
in three models: one for each predictor variable plus one that included both predictor
variables.  I then compared the three models to determine which, if any, was most
effective.

```{r}
# Create regression models for each predictor variable, as well as the combination of both.
tot_val_lm <- lm(occurrences ~ total_const_valuation, data = blocks_df)
one_unit_val_lm <- lm(occurrences ~ one_unit_valuation, data = blocks_df)
multi_lm <- lm(occurrences ~ total_const_valuation + one_unit_valuation, data = blocks_df)
# Plot the data with regression lines.
ggplot(data = blocks_df, aes(y = occurrences)) + geom_point(aes(x = total_const_valuation, color = "Total Valuation")) + 
  geom_smooth(aes(x = total_const_valuation, color="Total Valuation"), method = "lm", se=FALSE) +
  geom_point(aes(x = one_unit_valuation, color = "One Unit Valuation")) +
  geom_smooth(aes(x = one_unit_valuation, color="One Unit Valuation"), method = "lm", se=FALSE) +
  ggtitle("Occurrences vs. Valuation Data") + xlab("Valuation") + ylab("Occurrences") + 
  labs(color = "Legend") + scale_color_manual(values = c("Total Valuation" = "black", "One Unit Valuation" = "blue"))
```

# Analysis
I began by examining the summary output for the three regression models.  We can
see from the summary that using the total construction valuation as a predictor
creates a fairly weak model, with an R-squared value of about 0.375 and a significance
value slightly above 0.05.  The model using one unit construction valuation as the
predictor variable is slightly better, with an R-squared value of about 0.422 and
a significance value slightly below 0.05.  The model using both predictor variables
has a better fit than the other models, with an R-squared value of about 0.651, but
the significance values are slightly above the 0.05 level.

Next, I used the gvlma library to assess the model assumptions.  All models passed
the assessments, indicating that the models do not violate any of the basic assumptions
required for an effective regression model.  Next, I used the "deletion" function from
the same library to calculate deletion statistics for each model.  This calculation
indicates unusual values that may be having an undue influence on the model.  The
results indicate that deletion of the third row of data could improve model performance.
However, there is no obvious justification to remove that observation and doing
so may be manipulative.

```{r}
# View summaries of the three models.
summary(tot_val_lm)
summary(one_unit_val_lm)
summary(multi_lm)
# Assess model assumptions.
gvmodel_tot <- gvlma(tot_val_lm)
summary(gvmodel_tot)
gvmodel_one <- gvlma(one_unit_val_lm)
summary(gvmodel_one)
gvmodel_multi <- gvlma(multi_lm)
summary(gvmodel_multi)
# Deletion statistics for each model.
gvmodel_tot_del <- deletion.gvlma(gvmodel_tot)
summary(gvmodel_tot_del)
gvmodel_one_del <- deletion.gvlma(gvmodel_one)
summary(gvmodel_one_del)
gvmodel_multi_del <- deletion.gvlma(gvmodel_multi)
summary(gvmodel_multi_del)
```

# Implications
The results of this modeling indicate that a multiple regression model using yearly
total- and one-unit construction valuation as predictor variables could predict
the yearly observational occurrences of the Eastern Indigo snake in Florida.  The
R-squared value indicates that these variables could theoretically account for about
55% of the variation in yearly observational occurrences when applied as a general
model.

# Limitations
The most obvious limitation of this analysis is the very small sample size of the
observational occurrences data set.  Five year time blocks had as little as 12
occurrences, and aggregating the data this way resulted in only 10 observations.
The data itself has many potential issues as well.  It is based on human observation
and identification of species, often by non-experts.

# Conclusion
The data available regarding the Eastern Indigo snake is, not surprisingly, very
limited.  Species sighting data is relatively laborious to create, especially if
the curators of the data are attempting to verify the species identification and
other pertinent information.  After removing all of the observations that had issues
or were otherwise not usable, I found that the sample size had fallen well below
what I would consider sufficient for meaningful analysis.  That being said, there
were some statistically significance patterns between species occurrences and construction
permitting data, suggesting that there could be legitimacy to my hypothesis. This
was a very interesting and educational project, and I hope to continue this avenue
of research in the future.

# References
